---
title: 'WLE 650: Lab 1'
author: "Lizbeth G Amador"
date: "2024-01-19"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Your Assignment

In this lab, you will use the R CheatSheet and any online resources to:

1.  create an RMarkdown file that will be knitted as a word document you turn in
Please be sure to show all code and results in your knitted word document. Also, create sub-headings for tasks 2-5 above.  

### 2.  
Read in your multivariate data set.  

```{r, data upload}
#Setting directory
root_dir = "G:/Shared drives/MultivariateStatistics"
multivar_stud_data = "Data/StudentDataSets"

#read in phenology data
neonnpn = read.csv(file.path(root_dir, multivar_stud_data, "AmadorL_NeonNpn_OpenFLowers_conus.CSV"))
#remove extra columns
require(tidyverse, quietly = TRUE)
phe = neonnpn %>%
  select(-c("elevation_in_metersStat", "phenophase_status", "intensity_category_id", "intensity_value", "abundance_value", "elevation_in_meters", "first_yes_julian_date", "numdays_since_prior_no", "last_yes_julian_date", "numdays_until_next_no"))
```

### 3.  
Use the **head** function to show the columns and first several rows of your dataset.  
```{r}
head(phe, 6)
```

### 4  
Create a bulleted list describing the columns in your dataset and the type of variable that each column represents (e.g., ordinal, categorical, continuous, etc. per the lab lecture slides)  
* `observation_id`: unique observation identifier, integer  
* `update_datatime`: data (yyyy-MM-DD) and time (hh:mm:ss), character  
* `site_id`: unique identifier of site, integer  
* `genus`: organism's genus level, character  
* `species`: organism's species level, character  
* `common_name`: organism's common name, character  
* `kingdom`: organism's kingdom level, character  
* `individual_id`: unique identifier for each individual plant surveyed, integer  
* `phenophase_id`: unique identifier of phenophase groups, integer  
* `phenophase_description`: description of the phenophase, character  
* `observation_date`: Date (yyyy-mm-dd) of when an observation was originally made, character  
* `day_of_year`: converted observation data into a stadardised data measurement, integer  
* `first_yes_year`: year of when phenophase was first oberserved, integer  
* `first_yes_month`: Month of when phenophase was first oberserved, integer  
* `first_yes_day`: Day of when phenophase was first oberserved, integer  
* `first_yes_doy`: Day of year of when phenophase was first oberserved for a given site, integer  
* `first_yes_sample_size`:  Sample size of individuals that first flowered at a given site, integer  
* `mean_first_yes_doy`: Average day of year for first flowering individuals at a given site, integer  
* `data_name`: Dataset origin (NEON or NPN), character  
* `DomainID`: Unique identifier for NEON domains  
* `longitude`: Decimal degree longitude for observation point, numerical  
* `latitude`: Decimal degree latitude for observation point, numerical  

### 5.  
Create a plot of some of your data.  
```{r}
plot(phe$longitude, phe$latitude)

phe %>%
  ggplot() + geom_point(mapping = aes(x=longitude, y = latitude, color = data_name))
```



