---
title: 'WLE 650: Lab 1'
author: "Lizbeth G Amador"
date: "2024-01-19"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Your Assignment

In this lab, you will use the R CheatSheet and any online resources to:

### 1.

Create an RMarkdown file that will be knitted as a word document you turn in. Please be sure to show all code and results in your knitted word document. Also, create sub-headings for tasks 2-5 below.

### 2.

Read in your multivariate data set.

```{r, libraries, warning=FALSE}
require(tidyverse, quietly = TRUE)
```

```{r, data upload}
#Setting directory
root_dir = "G:/Shared drives/MultivariateStatistics"
multivar_stud_data = "Data/StudentDataSets"

#read in phenology data
neonnpn = read.csv(file.path(root_dir, multivar_stud_data, "AmadorL_NeonNpn_OpenFLowers_conus.CSV"))

#remove extra columns
phe = neonnpn %>%
  select(-c("elevation_in_metersStat", "phenophase_status", "intensity_category_id", "intensity_value", "abundance_value", "elevation_in_meters", "first_yes_julian_date", "numdays_since_prior_no", "last_yes_julian_date", "numdays_until_next_no"))
```

### 3.

Use the **head** function to show the columns and first several rows of your dataset.

```{r}
head(phe, 6)
```

### 4.

Create a bulleted list describing the columns in your dataset and the type of variable that each column represents (e.g., ordinal, categorical, continuous, etc. per the lab lecture slides)

-   `observation_id` : Unique observation identifier, integer
-   `update_datatime` : Data (yyyy-MM-DD) and time (hh:mm:ss), character
-   `site_id` : Unique identifier of site, integer
-   `genus` : Organism's genus level, character
-   `species` : Organism's species level, character
-   `common_name` : Organism's common name, character
-   `kingdom` : Organism's kingdom level, character
-   `individual_id` : Unique identifier for each individual plant surveyed, integer
-   `phenophase_id` : Unique identifier of phenophase groups, integer
-   `phenophase_description` : Description of the phenophase, character
-   `observation_date` : Date (yyyy-mm-dd) of when an observation was originally made, character
-   `day_of_year` : converted observation data into a stadardised data measurement, integer
-   `first_yes_year` : Year of when phenophase was first oberserved, integer
-   `first_yes_month`: Month of when phenophase was first oberserved, integer
-   `first_yes_day` : Day of when phenophase was first oberserved, integer
-   `first_yes_doy` : Day of year of when phenophase was first oberserved for a given site, integer
-   `first_yes_sample_size` : Sample size of individuals that first flowered at a given site, integer
-   `mean_first_yes_doy` : Average day of year for first flowering individuals at a given site, integer
-   `data_name` : Dataset origin (NEON or NPN), character
-   `DomainID` : Unique identifier for NEON domains
-   `longitude` : Decimal degree longitude for observation point, numerical
-   `latitude` : Decimal degree latitude for observation point, numerical

### 5.

Create a plot of some of your data.

```{r, fig1, fig.height = 5, fig.width = 6, fig.align = "center"}
phe %>%
  ggplot() + 
  geom_point(mapping = aes(x=longitude, y = latitude, color = data_name)) + 
  xlab("Longitude") + 
  ylab("Latitude") +
  scale_color_manual(name="Dataset", values = alpha(c("hotpink", "darkgrey"), 
                      c(0.9, 0.35))) + 
  theme(panel.background = element_rect(fill='transparent'),
        legend.title=element_text(face="bold"),
        legend.key = element_rect(fill = "white"),
        legend.position = c(0.9, 0.2), 
        legend.background = element_rect(size=0.5, linetype="solid", colour ="black"),
        axis.line = element_line(color = "black",linewidth = 0.5, linetype = 1))
```

```{r, fig2, fig.height = 5, fig.width = 7, fig.align = "center"}
par(mfrow=c(1,2))
#All FFD
hist(phe$first_yes_doy, main = "First Flower Date (FFD)",
     xlab="Day of Year (DOY)")
#All avg FFD
hist(phe$mean_first_yes_doy, main = "Average First Flower Date (FFD)",
     xlab="Day of Year (DOY)")
```
